---
title: "External gas calibration (ExtGasCal) workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow_ExtGasCal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

External gas calibration is a new approach used in combination with ICP techniques 
and has been applied for the determination of microplastics with [single particle 
ICP-MS](https://link.springer.com/article/10.1007/s00216-025-05934-9) and [ETV/ICP-MS](paper link). The working principle is based on the 
introduction of a dynamically diluted calibration gas into the ICP-MS system. 
Consequently, data processing requires calculation of a mean value from the 
continuous signal curves.

The workflow external gas calibration (ExtGasCal) is compatible with ICP-MS or 
-OES data.

In the package *ETVapp*, we provide the example data of [published microplastic measurements](https-Link). They can be accessed by `ETVapp::ETVapp_testdata`.

## Calibration data

Import the calibration files through the following code.

```{r cali_import}
library(ETVapp)
cali_imp <- ETVapp::ETVapp_testdata[['ExtGasCal']][['Cali']]
head(cali_imp[[1]])
```

Select the analyte (parameter *c1*) and set a filter length *fl* for 
Savitzky-Golay smoothing. Set *fl* to `NULL` to omit the step. `Process_data()` 
provides optional correction with an internal standard (parameter *c2*).

```{r cali_processing}
anlt <- "13C"
time_col <- "Time"
cali_pro <- lapply(cali_imp, function(x) {
  process_data(data = x[,c(time_col, anlt)], c1 = anlt, fl = 9, wf = "ExtGasCal")
})
head(cali_pro[[1]])
```

### Signal evaluation

Receive mean signal intensities *via* the peak picking method "mean_signal". The 
signal integration range is defined by *peak_start* and *peak_end*. Gas 
flows [mL/min] are asigned to the signal data with `tab_cali()`. The parameter 
*fac* enables the input of a conversion factor, *e.g.* for mass percentage, molar 
fractions, when computing the gas flows in µg/s.

```{r get_cali_peaks}
cali_sig <- ldply_base(1:length(cali_pro), function(i) {
  get_peakdata(
    cali_pro[[i]], 
    int_col = anlt, 
    PPmethod = "mean_signal",
    peak_start = 78, 
    peak_end = 96
  )
})

fac <- 1.661 * 0.01104347 * 12/44
cali_peaks <- tab_cali(peak_data = cali_sig, wf = "ExtGasCal", gas_flow = seq(0,50,length.out=nrow(cali_sig)), fac = fac)
print(cali_peaks)
```

The following code will plot the time scans of the calibration standards. The 
evaluated signal range is framed by grey vertical lines.

```{r cali_peak_detection_plot, fig.width=3*3, fig.height=2*3, out.width='100%', fig.align='center'}
par(mfrow=c(3,3))
for (i in 1:9) {
  plot(cali_pro[[i]], type="l", ylab = "Intensity [cps]", xlab = "Time [s]", main = paste(cali_peaks[i,5], "mL/min"))
  abline(v=cali_peaks[i,2:3], col=grey(0.8))
}
par(mfrow=c(1,1))
```

Peak areas are plotted against the gas flows. Calibration parameter obtained 
through linear regression are provided as output *data.frame*.

```{r calibration_model, fig.width=3, fig.height=3}
(cm <- calc_cali_mod(df = cali_peaks[,c(5,4)], wf = "ExtGasCal"))
plot(cali_peaks[,c(5,4)])
abline(a = cm[1,3], b = cm[1,1])
```

## Sample analysis

Repeat the procedure for the sample data.

```{r load_sample}
smpl_imp <- ETVapp::ETVapp_testdata[['ExtGasCal']][['Samples']]

smpl_pro <- lapply(smpl_imp, function(x) {
  process_data(data = x, c1 = anlt, fl = 9, wf = "ExtGasCal")
})
head(smpl_pro[[1]])

ps<- 90
pe <- 118
cf <- 50

smpl_peaks <- ldply_base(1:length(smpl_pro), function(i) {
  get_peakdata(
    smpl_pro[[i]], 
    int_col = anlt, 
    PPmethod = "Peak (manual)",
    minpeakheight = 1000000,
    peak_start = ps, 
    peak_end = pe
  )
})
print(smpl_peaks)
```
Generate baseline data from a time scan excerpt and plot the sample time scan with integration parameter.

```{r get_sample_BL}
smpl_BL <- lapply(1:length(smpl_pro), function(i) {
  flt <- (min(which(smpl_pro[[i]][,time_col]>=ps))-cf):(max(which(smpl_pro[[i]][,time_col]<=pe))+cf)
  ETVapp:::blcorr_col(
    df = smpl_pro[[i]][flt,c(time_col, anlt)],
    nm = anlt, 
    BLmethod = "modpolyfit",
    rval = "baseline", 
    amend = "_BL")
})
```

```{r sample_peak_detection_plot, fig.width=3*3, fig.height=2*2, out.width='100%', fig.align='center'}
par(mfrow=c(1,2))
for (i in 1:2) {
  plot(smpl_pro[[i]], type="l", ylab = "Intensity [cps]", xlab = "Time [s]")
  lines(x = smpl_BL[[i]][,c("Time")], y = smpl_BL[[i]][,3], col = "blue")
  abline(v=smpl_peaks[i,2:3], col=grey(0.8))
}
par(mfrow=c(1,1))
```

Compute the sample result based on calibration model with the following code. 
The input of a mass fraction allows for result calculation of the micro plastic 
content (content as analyte) from the carbon mass (content as element). 

```{r tab_sample_result}
w_PE <- 0.856
sample_mass = c(0.99, 1.02)

t(tab_result(
  smpl_peaks, 
  a = cm[1,3], 
  b = cm[1,1], 
  wf = "ExtGasCal",
  mass_fraction2 = w_PE, 
  sample_mass = sample_mass
))
```

## Limits of detection and quantification

Blank measurements are imported and processed as follows. For estimating the 
limit of detection (LOD) and quantification (LOQ) according to the blank value 
method, integrate the blank signals in the peak time window. 

```{r load_blanks, fig.width=2*4, fig.height=3*2, out.width='100%', fig.align='center'}
blnk_imp <- ETVapp::ETVapp_testdata[['ExtGasCal']][['Blanks']]
blnk_pro <- lapply(1:length(blnk_imp), function(i) {
  process_data(data = blnk_imp[[i]], c1 = anlt, fl = 9, wf = "ExtGasCal")
})
blnk_BL <- lapply(1:length(blnk_pro), function(i) {
  flt <- (min(which(blnk_pro[[i]][,time_col]>=ps))-cf):(max(which(blnk_pro[[i]][,time_col]<=pe))+cf)
  ETVapp:::blcorr_col(
    df = blnk_pro[[i]][flt,c(time_col, anlt)],
    nm = anlt, 
    BLmethod = "modpolyfit",
    rval = "baseline", 
    amend = "_BL")
})

blnk_peaks <- ldply_base(1:length(blnk_pro), function(i) {
  get_peakdata(
    blnk_pro[[i]], 
    int_col = anlt, 
    PPmethod = "Peak (manual)",
    BLmethod = "modpolyfit",
    peak_start = ps, 
    peak_end = pe
  )
})

print(blnk_peaks)

par(mfrow=c(3,3))
for (i in 1:min(length(blnk_pro), 10)) {
  ylim <- c(0, max(sapply(blnk_pro, function(x) {max(x[,2])})))
  plot(blnk_pro[[i]], type="l", main = names(blnk_pro)[i], ylab="Intensity [cps]", xlab = "Time [s]",ylim=ylim)
  lines(x = blnk_BL[[i]][,c("Time")], y = blnk_BL[[i]][,3], col = "blue")
  abline(v=blnk_peaks[i,2:3], col=grey(0.8))
}
par(mfrow=c(1,1))
```

The LOD and LOQ are estimated as three and ten times the standard deviation of 
the blank values divided by the slope of the linear calibration curve. At least 
three input values are required for statistical evaluation. Less than the 
recommended 10 entries will still give a result. The input of a theoretical 
sample mass will compute the LOD and LOQ per sample mass. Figures of merit are 
provided related to carbon (measured element) and polyethylene (analyte).

```{r get_LOX}
t(tab_LOX(x = blnk_peaks[,4], cali_slope = cm[1,1], wf = "ExtGasCal", mass_fraction2 = w_PE))
```
